Mainstreaming Scientific Racism

### 2022-10-31 talk/talk.qmd
working on slides

---
format: 
    revealjs:
        scrollable: true
        width: 1440
        height: 1080
        toc: true
        slide-number: true
        theme: [default, custom.scss]

title: "Mainstreaming Scientific Racism"
author: "Dan Hicks & Emilio Lobato"
institute: "UC Merced"
---

@@

@@

:::: columns 
::: {.column width="30%"}
![[Charles Davenport (1866-1944)](https://en.wikipedia.org/wiki/Charles_Davenport), circa 1929. [Source](https://en.wikipedia.org/wiki/File:Charles_Benedict_Davenport.jpg)](img/Charles_Benedict_Davenport.jpg){width="100%"}
:::

::: {.column width="30%"}
![[Franz Boas (1858-1942)](https://en.wikipedia.org/wiki/Franz_Boas), circa 1915. [Source](https://en.wikipedia.org/wiki/File:FranzBoas.jpg)](img/FranzBoas.jpg)
:::

::: {.column width="30%"}
![[Margaret Mead (1901-1978)](https://en.wikipedia.org/wiki/Margaret_Mead), 1948. [Source](https://en.wikipedia.org/wiki/File:Margaret_Mead_(1901-1978).jpg)](img/Margaret_Mead_(1901-1978).jpg)
:::
::::

- 1910: Davenport founds Eugenics Records Office at Cold Spring Harbor
- 1926: NRC "Committee on the Negro" includes both Davenport and Boas: "neither was influential enough to veto his rival's participation" [@BarkanRetreatScientificRacism1992 113]
- 1930s: *Coming of Age in Samoa* *[cite]* popularizes cultural anthropology, "free[ing] anthropology from the shackles of biology" [@BarkanRetreatScientificRacism1992 134]
- 1940: Carnegie Institute defunds Eugenics Records Office
- 1950: "For all practical social purposes 'race' is not so much a biological phenomenon as a social myth." *[https://unesdoc.unesco.org/ark:/48223/pf0000128291]*

## Pioneer Fund

*[todo]*

@@

> Ultimately, anti-'racial' science played a defining role in the Supreme Court’s dismissal of the 'separate-but-equal' principle in education .... [@SchafferScientificRacismAgain2007 261]

@@

> A scientific approach to racial questions, [free from sentiment, propaganda or bias]{.blue}, may lead to a more reasonable attitude to the problems which agitate considerable sections of mankind. This need for a calmer approach to problems of race on the part of [civilised peoples as well as more primitive tribes]{.red} makes it desirable that a journal should exist in which the problems of racial origin and racial relationships can be quietly discussed, [without rancour or bigotry and with the primary aim of elucidating facts]{.blue}. (Gates, *MQ* 1:1 [1960], 11)

# Race science agntology

## The Tobacco Strategy

@@

[Did Pioneer and *Mankind Quarterly* function in this way for race science?]{.blue}

# Data and methods

## Did PF and MQ function as a race science echo chamber?

@@

@@

::::: columns
:::: column
::: {.smaller}
| | |
|:-----|:-----|
| [Thomas J. Bouchard, Jr.](https://en.wikipedia.org/wiki/Thomas_J._Bouchard_Jr.) | psychology |
| Brunetto Chiarelli | | anthropology? |
| [Hans Eysenck](https://en.wikipedia.org/wiki/Hans_Eysenck) | psychology |
| [Robert Gordon](https://en.wikipedia.org/wiki/Robert_A._Gordon) | sociology |
| [Linda Gottfredson](https://en.wikipedia.org/wiki/Linda_Gottfredson) | psychology |
| [Garrett Hardin](https://en.wikipedia.org/wiki/Garrett_Hardin) | ecology |
| [Joseph M. Horn](https://en.wikipedia.org/wiki/Joseph_M._Horn) | psychology |
| [Lloyd Humphreys](https://en.wikipedia.org/wiki/Lloyd_Humphreys) | psychology |
| [Arthur Jensen](https://en.wikipedia.org/wiki/Arthur_Jensen) | psychology |
| [Michael Levin](https://en.wikipedia.org/wiki/Michael_Levin) | philosophy |
| [Richard Lynn](https://en.wikipedia.org/wiki/Richard_Lynn) | psychology |
| [R. Travis Osborne](https://en.wikipedia.org/wiki/R._Travis_Osborne) | psychology |
| [J. Phillippe Rushton](https://en.wikipedia.org/wiki/J._Phillippe_Rushton) | psychology |
| [Audrey M. Shuey](https://en.wikipedia.org/wiki/Audrey_M._Shuey) | psychology |
| [Philip A. Vernon](https://en.wikipedia.org/wiki/Philip_A._Vernon) | psychology |
| [Daniel Vining, Jr.](https://en.wikipedia.org/wiki/Daniel_Vining_Jr.) | demography |
:::
::::

@@

## Corpus

|                                |       |
|:-------------------------------|------:|
| Behavior Genetics              |  2268 |
| Behavioral & Brain Sciences    |   898 |
| Intelligence                   |  1237 |
| **Mankind Quarterly**          |  1821 |
| Person. & Individ. Differences |  7274 |
| Psychological Reports          | 21398 |

@@

\
- 3 journals published by American Psychological Association not available
- PID is missing many relevant book reviews and essays
	- eg, [Rushton's review of Lynn's book on Pioneer](https://www.sciencedirect.com/science/article/pii/S0191886901001337) *[confirm not included]*

@@

![Diachronic evolution of topic categories in *Philosophy of Science*, 1934–2015. @MalaterreWhatThisThing2019 fig. 2.](img/MalaterreWhatThisThing2019.jpeg){width="80%"}

- Statistical method for clustering documents and words simultaneously
- Documents are composed of topics; topics are composed of words
- Custom R package [`tmfast`](https://github.com/dhicks/tmfast)

<!-- 
## Gould's critique of factor analysis

@@

## Topic modeling as phenomena construction

@@

These words tend to be used together; **but why?**\
These test items tend to be answered in the same way; **but why?**
 -->

# Mainstreaming race science

## Umap interactive

```{r}
#| out-width: 1440px
#| out-height: 1080px
#| fig-cap: "UMAP visualization of Hellinger similarity, medium vocabulary, k = 30"
library(readr)
library(here)
library(plotly)

vis_dir = here('out')

>>>>

---
format: 
    revealjs:
        scrollable: true
        width: 1.1*1440
        height: 1080
        toc: true
        toc-depth: 1
        toc-title: Overview
        slide-number: true
        theme: [default, custom.scss]

title: "Mainstreaming Scientific Racism"
author: "Dan Hicks & Emilio Lobato"
institute: "University of California, Merced"
bibliography: ../paper/race_science.yaml
---

@@

```{r setup}
library(readr)
library(here)
library(plotly)
library(gt)

vis_dir = here('out')
```

@@

:::: columns 
::: {.column width="33%"}
![[Charles Davenport (1866-1944)](https://en.wikipedia.org/wiki/Charles_Davenport), circa 1929. [Source](https://en.wikipedia.org/wiki/File:Charles_Benedict_Davenport.jpg)](img/Charles_Benedict_Davenport.jpg){width="70%"}
:::

::: {.column width="33%"}
![[Franz Boas (1858-1942)](https://en.wikipedia.org/wiki/Franz_Boas), circa 1915. [Source](https://en.wikipedia.org/wiki/File:FranzBoas.jpg)](img/FranzBoas.jpg){width="70%"}
:::

::: {.column width="33%"}
![[Margaret Mead (1901-1978)](https://en.wikipedia.org/wiki/Margaret_Mead), 1948. [Source](https://en.wikipedia.org/wiki/File:Margaret_Mead_(1901-1978).jpg)](img/Margaret_Mead_(1901-1978).jpg){width="70%"}
:::
::::

- 1910: Davenport founds Eugenics Records Office at Cold Spring Harbor
- 1926: NRC "Committee on the Negro" includes both Davenport and Boas: "neither was influential enough to veto his rival's participation" [@BarkanRetreatScientificRacism1992 113]
- 1930s: *Coming of Age in Samoa* [@MeadComingAgeSamoa1928] popularizes cultural anthropology, "free[ing] anthropology from the shackles of biology" [@BarkanRetreatScientificRacism1992 134]
- 1940: Carnegie Institute defunds Eugenics Records Office
- 1950: "For all practical social purposes 'race' is not so much a biological phenomenon as a social myth" [@BeagleholeRaceQuestion1950; though see @BrattainRaceRacismAntiracism2007]

## The decline of eugenics and the Pioneer Fund {.smaller}

:::: columns
::: {.column width="30%"}
![[Wickliffe Draper (1891-1972)](https://en.wikipedia.org/wiki/Wickliffe_Preston_Draper), textile heir, eugenicist, segregationist, 1918. [Source](https://en.wikipedia.org/wiki/File:Wickliffe_Draper.jpg)](img/Draper.jpg)
:::

::: {.column width="30%"}
![[Harry Laughlin (1880-1943)](https://en.wikipedia.org/wiki/Harry_Laughlin), 
superintendent of Eugenics Records Office, anti-immigrationist, c. 1929. [Source](https://en.wikipedia.org/wiki/File:Harry_Laughlin.jpg)](img/Harry_Laughlin.jpg){width="100%"}
:::

::: {.column width="30%"}
![[Frederick Osborn (1889-1981)](https://en.wikipedia.org/wiki/Frederick_Osborn), railroad heir, eugenicist, shaped postwar pivot to "positive eugenics," undated. [Source](https://www.meehanreports.com/eugen-PT-pt1.htm)](img/osbornphoto.jpg){width="100%"}
:::
::::

Founded in 1937 to "support academic research and the \'[sic]dissemination of information, into the 'problem of heredity and eugenics' and 'the problems of race betterment'" [@MehlerFoundationFascismNew1989 21, quoting Laughlin]
<!-- 
- Presidents
	- Harry Laughlin, 1937-1941
	- Frederick Osborn, 1941–1958
	- Harry Weyher, Jr., 1958-2002
	- J. Phillippe Rushton, 2002-12
	- Richard Lynn, 2012-??
 -->

@@

> Ultimately, anti-'racial' science played a defining role in the Supreme Court’s dismissal of the 'separate-but-equal' principle in education .... [@SchafferScientificRacismAgain2007 261; see also @JacksonScienceSegregationRace2005]

@@

> [A scientific approach to racial questions, free from sentiment, propaganda or bias]{.blue}, may lead to a more reasonable attitude to the problems which agitate considerable sections of mankind. This need for a calmer approach to problems of race on the part of [civilised peoples as well as more primitive tribes]{.red} makes it desirable that a journal should exist in which the problems of racial origin and racial relationships can be quietly discussed, [without rancour or bigotry and with the primary aim of elucidating facts]{.blue}. (Gates, *MQ* 1:1 [1960], 11)

# Race science agnotology

## The Tobacco Strategy {.smallish}

@@

[Did Pioneer and *Mankind Quarterly* function in this way for race science?]{.red}

<!-- 
## PF and MQ as a race science echo chamber?

@@

\
 -->

@@

::::: columns
:::: column
::: {.smaller}
| | |
|:-----|:-----|
| [Thomas J. Bouchard, Jr.](https://en.wikipedia.org/wiki/Thomas_J._Bouchard_Jr.) | psychology |
| Brunetto Chiarelli | anthropology? |
| [Hans Eysenck](https://en.wikipedia.org/wiki/Hans_Eysenck) | psychology |
| [Robert Gordon](https://en.wikipedia.org/wiki/Robert_A._Gordon) | sociology |
| [Linda Gottfredson](https://en.wikipedia.org/wiki/Linda_Gottfredson) | psychology |
| [Garrett Hardin](https://en.wikipedia.org/wiki/Garrett_Hardin) | ecology |
| [Joseph M. Horn](https://en.wikipedia.org/wiki/Joseph_M._Horn) | psychology |
| [Lloyd Humphreys](https://en.wikipedia.org/wiki/Lloyd_Humphreys) | psychology |
| [Arthur Jensen](https://en.wikipedia.org/wiki/Arthur_Jensen) | psychology |
| [Michael Levin](https://en.wikipedia.org/wiki/Michael_Levin) | philosophy |
| [Richard Lynn](https://en.wikipedia.org/wiki/Richard_Lynn) | psychology |
| [R. Travis Osborne](https://en.wikipedia.org/wiki/R._Travis_Osborne) | psychology |
| [J. Phillippe Rushton](https://en.wikipedia.org/wiki/J._Phillippe_Rushton) | psychology |
| [Audrey M. Shuey](https://en.wikipedia.org/wiki/Audrey_M._Shuey) | psychology |
| [Philip A. Vernon](https://en.wikipedia.org/wiki/Philip_A._Vernon) | psychology |
| [Daniel Vining, Jr.](https://en.wikipedia.org/wiki/Daniel_Vining_Jr.) | demography |
:::
::::

@@

![Journals publishing 6 or more Pioneer-funded researchers, WOS](img/wos_results.png)

## Corpus {.smallish}

|                                |       |
|:-------------------------------|------:|
| Behavior Genetics              |  2268 |
| Behavioral & Brain Sciences    |   898 |
| Intelligence                   |  1237 |
| **Mankind Quarterly**          |  1821 |
| Personality & Individual<br> Differences |  7274 |
| Psychological Reports          | 21398 |

@@

- 3 journals published by American Psychological Association not available
- In B&BS, focal papers + commentaries all bundled together
- PID is missing many relevant book reviews and essays
	- eg, [Rushton's review of Lynn's book on Pioneer](https://www.sciencedirect.com/science/article/pii/S0191886901001337)

@@

:::: columns
::: {.column width="60%"}
![Diachronic evolution of topic categories in *Philosophy of Science*, 1934–2015. @MalaterreWhatThisThing2019 fig. 2.](img/MalaterreWhatThisThing2019.jpeg){width="100%"}
:::

::: {.column width="40%"}
- Statistical method for clustering documents and words simultaneously
- [Can we identify a race science topic?]{.red}
- [Does this topic start in MQ, spread to mainstream journals?]{.red}
:::
::::

::: {.smaller}
- Custom R package [`tmfast`](https://github.com/dhicks/tmfast)
- 18 different models (3 vocabulary sizes $\times$ $k$ = 5, 10, 20, ..., 50)
:::

## UMAP interactive

```{r}
#| out-width: 1.1*1440px
#| out-height: 1080px
#| fig-cap: "UMAP visualization of Hellinger similarity, medium vocabulary, k = 30"

## 3 focal topics {.smaller}

```{r}
beta_gt = read_rds(here(vis_dir, '??_beta_table.Rds'))
beta_gt
```

---

*[focal topics by journal over time]*

# Conclusion

- Race & intelligence research did not appear to originate in MQ and spread to other journals
- Instead, Intelligence and PID were the primary "echo chambers" used to legitimize this psychological form of race science
- "The agnotology is coming from inside the house"

## Different topics or different communities?  {visibility="uncounted"}

![Authors by topic](../out/_authors.png)

## Gould's critique of factor analysis {visibility="uncounted"}

@@

## Topic modeling as phenomena construction {visibility="uncounted"}

@@

These words tend to be used together; **but why?**\
These test items tend to be answered in the same way; **but why?**

## References {visibility="uncounted"}
